Lazy Linearity for a Core Functional Language

### 2025-10-14 popl26.tex
.

\begin{abstract}
Traditionally, \ROUNDTWO{in linearly typed languages, consuming} a linear resource is synonymous
with its syntactic occurrence in the program. However, under the lens of
non-strict evaluation, linearity can be further understood semantically,
where a syntactic occurrence of a resource does not necessarily entail
using that resource when the program is executed.
%
\ROUNDTWO{While this distinction is largely unexplored, it becomes unavoidable
    for} Haskell's optimising compiler, which heavily rewrites the source
    program in ways that break syntactic linearity but preserve the program's
    semantics.
% \ROUNDTWO{This distinction becomes vital in the Haskell optimising compiler,} where
% optimisations heavily rewrite the source program in ways that break syntactic
% linearity but preserve the program's semantics
%
% This distinction is 
% \ROUNDTWO{This distinction is critical for the Haskell optimising compiler, 
%
% when considering the internal language of an optimising compiler, where
% optimisations heavily rewrite the source program in ways that break syntactic
% linearity but preserve the program's semantics.
%
%
% %
% Semantic linearity is especially necessary in optimising compilers for
% languages combining linearity and laziness: optimisations leverage laziness to
% heavily rewrite the source program, pushing the interaction of linearity and
% laziness to its limit, regardless of the original program typing linearity
% conservatively.
%
%
We introduce Linear Core, \ROUNDTWO{a novel system which accepts the lazy
semantics of linearity statically} and
% \ROUNDTWO{Linear Core explores the interaction between laziness and linearity}
is suitable for the Core intermediate language of the
Glasgow Haskell Compiler. We prove \ROUNDTWO{that} Linear Core is sound,
guarantees linear resource usage, and \ROUNDTWO{that} multiple
optimising transformations preserve linearity in Linear Core while failing
to do so in Core. We have implemented Linear Core as a compiler plugin to
validate the system against linearity-heavy libraries, including
\texttt{linear-base}.%^, in the heart of the compiler.
\end{abstract}

>>>>

\begin{abstract}
Traditionally, \ROUNDTWO{in linearly typed languages, consuming} a linear resource is synonymous
with its syntactic occurrence in the program. However, under the lens of
non-strict evaluation, linearity can be further understood semantically,
where a syntactic occurrence of a resource does not necessarily entail
using that resource when the program is executed.
%
\ROUNDTWO{While this distinction has been largely unexplored, it turns
  out to be inescapable in} Haskell's optimising compiler, which heavily rewrites the source
    program in ways that break syntactic linearity but preserve the program's
    semantics.
% \ROUNDTWO{This distinction becomes vital in the Haskell optimising compiler,} where
% optimisations heavily rewrite the source program in ways that break syntactic
% linearity but preserve the program's semantics
%
% This distinction is 
% \ROUNDTWO{This distinction is critical for the Haskell optimising compiler, 
%
% when considering the internal language of an optimising compiler, where
% optimisations heavily rewrite the source program in ways that break syntactic
% linearity but preserve the program's semantics.
%
%
% %
% Semantic linearity is especially necessary in optimising compilers for
% languages combining linearity and laziness: optimisations leverage laziness to
% heavily rewrite the source program, pushing the interaction of linearity and
% laziness to its limit, regardless of the original program typing linearity
% conservatively.
%
%
We introduce Linear Core, \ROUNDTWO{a novel system which accepts the lazy
semantics of linearity statically} and
% \ROUNDTWO{Linear Core explores the interaction between laziness and linearity}
is suitable for \ROUNDTWO{lazy languages such as } the Core intermediate language of the
Glasgow Haskell Compiler. We prove \ROUNDTWO{that} Linear Core is sound,
\ROUNDTWO{guaranteeing} linear resource usage, and \ROUNDTWO{that} multiple
optimising transformations preserve linearity in Linear Core while failing
to do so in Core. We have implemented Linear Core as a compiler plugin to
validate the system against linearity-heavy libraries, including
\texttt{linear-base}.%^, in the heart of the compiler.
\end{abstract}
